Ethereum: A Secure Decentralised Generalised Transaction Ledger

### 2014-04-19 Paper.tex
Modified Merkle Patricia Trie.

Dealings in this proposed system would have several attributes not often found in the real word. The incorruptibility of judgement, often difficult to find, comes naturally from a disinterested algorithmic interpreter. Transparency, through being about to see exactly how a state or judgement came about through the transaction log and rules or instructional codes never happens perfectly in human-based systems.

Overall, we wish to provide a system such that the user can be guaranteed that any other individuals, systems or organisations that they interact with, they can do so with absolute confidence in the possible outcomes and how they might come about.

%\subsection{Previous Work} \label{ch:previous}
%E language. Smart contracts. Bitcoin, namecoin \&c. Mastercoin.

@@

Ethereum, taken as a whole, can be viewed as a transaction-based state machine: we begin with a genesis state and incrementally execute transactions to morph it into some final state. It is this final state which we accept as the canonical ``version'' of the world of Ethereum. The state can include such information as account balances, reputations, trust arrangements, data pertaining to information of the physical world; in short, anything that can currently be represented by a computer is admissable. Transactions thus represent a valid arc between two states; the `valid' part is important---there exist far more invalid state changes than valid state changes. Invalid state changes might, \eg be things such as reducing an account balance without an equal and opposite increase elsewhere. We define a valid state transitions as one which comes about through a transaction. Formally:

@@

Transactions are collated into blocks; we then chain blocks together using a cryptographic hash as a means of reference. Blocks function as a journal, recording a series of transactions together with the previous block and an identifier for the final state (though do not store the final state itself---that would be far too big). They also punctuate the transaction series with incentives for nodes to \textit{mine}. This incentivisation takes places as a state-transition function, adding value to a nominated account.

@@

In order to incentivise computation within the network, there needs to be an agreed method for transmitting value. To address this issue, Ethereum has an intrinsic currency, Ether, known also as {\small ETH}. The smallest subdenomination of Ether, and thus the one in which all integer values of the currency are counted, is the Wei. One Ether is defined as being $10^{18}$ Wei. There exist other subdenominations of Ether:
\par`
\begin{center}
\begin{tabular}{rl}
\toprule
Multiplier & Name \\
\midrule
$10^0$ & Wei \\
$10^{12}$ & Szabo \\
$10^{15}$ & Finney \\
$10^{18}$ & Ether \\
\bottomrule
\end{tabular}
\end{center}
\par

@@

We define the rollback state $\mathbb{S}_0$:
\begin{eqnarray}
\mathbb{S}_0 & \equiv & \mathbb{S} \quad \text{except:} \\
\mathbb{S}_0[S(T)]_b & \equiv & \mathbb{S}[S(T)]_b - v_0 \\
\mathbb{S}_0[S(T)]_n & \equiv & \mathbb{S}[S(T)]_n + 1
\end{eqnarray}

@@

We define $\mathbb{S}_1$, the rollback state as the original state but with the value transferred to the recipient:
\begin{eqnarray}
\mathbb{S}_1 & \equiv & \mathbb{S} \quad \text{except:} \\
\mathbb{S}_1[r]_b & \equiv & \mathbb{S}_1[r]_b + v
\end{eqnarray}

If the recipient account contains no code to be executed (i.e. it's not a contract), then we define the final state as being equivalent to the rollback state. In the case that the recipient contains code to be executed (i.e. that the account is a contract), then the contract's body code (identified as the fragment whose SHA3 hash is $\mathbb{S}[r]_c$) is executed according to the execution model (see section \ref{ch:model}). Just as with contract creation, if the execution halts due to an exhausted gas supply, then no gas is refunded to the caller and the state is reverted to the point immediately prior to code execution.
\begin{eqnarray}
\mathbb{S}_P & \equiv & \begin{cases}
\mathbb{S}_1 & \text{if} \quad \mathbb{S}_1[r]_c = \varnothing \\
\mathbb{S}_1 & \text{if} \quad \mathbb{S}^{**} = \varnothing \\
\mathbb{S}^{**} & \text{otherwise}
\end{cases} \\
(\mathbb{S}^{**}, g', \mathbf{o}) & \equiv & \Xi(\mathbb{S}_1, g, I) \\
I_a & \equiv & a \\
I_o & \equiv & o \\
I_p & \equiv & p \\
I_\mathbf{d} & \equiv & \mathbf{d} \\
I_s & \equiv & s \\
I_v & \equiv & v \\
\text{Let} \quad \mathtt{\tiny SHA3}(I_\mathbf{b}) & = & \mathbb{S}[r]_c
\end{eqnarray}

@@

State database won't keep everything; maintains an age and eventually throws out nodes that are not recent enough or checkpoints.

Nodes in state trie that haven't sent/received a transaction in over X blocks could be thrown out, reducing Ether-leakage and reducing the growth of the state database.

@@

This paper has introduced, discussed and formally defined the protocol of Ethereum.

@@

Vitalik Buterin first proposed in the original Ethereum whitepaper (first distributed in December 2013) moving from a suffocating state-transaction function with trivial state to something much more general and expansive. Useful corrections and suggestions were provided by a number of others from the Ethereum community including Aeron Buchanan, Nick Savers and Marko Simovic.

@@

\section{Recursive Length Prefix}\label{app:rlp}
This is a serialisation method for encoding arbitrarily structured binary data (byte arrays). Within RLP, integers are specified as the byte array equal to the big-endian representation of the value.

@@

@@

\begin{eqnarray}
\mathtt{\tiny HP}(\mathbf{x}, t) & \equiv & ...
\end{eqnarray}

\section{Modified Merkle Patricia Tree}\label{app:trie}
The modified Merkle Patricia tree (trie) provides a persistent data structure to map between arbitrary-length binary data (byte arrays). It is defined in terms of a mutable data structure to map between 256-bit binary fragments and arbitrary-length binary data, typically implemented as a database.

@@

\section{The State DB}\label{app:state}

The state database is a mapping between SHA3-256 hashes (i.e. 256-bit binary fragments) and arbitrary byte data.

@@

The method of signing transactions is similar to the `Electrum style signatures'; it utilises the SECP-256k1 curve. This has yet to be formally specified.

@@

\begin{tabular*}{\columnwidth}[h]{rlrrl}
\toprule
\multicolumn{5}{c}{\textbf{50s: Stack, Memory, Storage and Flow Operations}} \vspace{5pt} \\
\textbf{Value} & \textbf{Mnemonic} & $\delta$ & $\alpha$ & \textbf{Description} \vspace{5pt} \\
0x50 & {\small POP} & 1 & 0 & Remove item from stack. \\
\midrule
0x51 & {\small DUP} & 1 & 2 & Duplicate stack item. \\
&&&& $\mathbb{M}'_\mathbf{s}[0] \equiv \mathbb{M}_\mathbf{s}[0]$ \\
\midrule
0x52 & {\small SWAP} & 2 & 2 & Exchange stack items. \\
&&&& $\mathbb{M}'_\mathbf{s}[0] \equiv \mathbb{M}_\mathbf{s}[1]$ \\
&&&& $\mathbb{M}'_\mathbf{s}[1] \equiv \mathbb{M}_\mathbf{s}[0]$ \\
\midrule
0x53 & {\small MLOAD} & 1 & 1 & Load word from memory. \\
&&&& $\mathbb{M}'_\mathbf{s}[0] \equiv \mathbb{M}_\mathbf{m}[\mathbb{M}_\mathbf{s}[0] \dots (\mathbb{M}_\mathbf{s}[0] + 31) ]$ \\
&&&& $\mathbb{M}'_i \equiv \max(\mathbb{M}_i, \ceil{ (\mathbb{M}_\mathbf{s}[0] + 32) \div 32 })$ \\
\midrule
0x54 & {\small MSTORE} & 2 & 0 & Save word to memory. \\
&&&& $\mathbb{M}'_\mathbf{m}[ \mathbb{M}_\mathbf{s}[0] \dots (\mathbb{M}_\mathbf{s}[0] + 31) ] \equiv \mathbb{M}_\mathbf{s}[1]$ \\
&&&& $\mathbb{M}'_i \equiv \max(\mathbb{M}_i, \ceil{ (\mathbb{M}_\mathbf{s}[0] + 32) \div 32 })$ \\
\midrule
0x55 & {\small MSTORE8} & 2 & 0 & Save byte to memory. \\
&&&& $\mathbb{M}'_\mathbf{m}[ \mathbb{M}_\mathbf{s}[0] ] \equiv (\mathbb{M}_\mathbf{s}[1] \mod 256) $ \\
&&&& $\mathbb{M}'_i \equiv \max(\mathbb{M}_i, \ceil{ (\mathbb{M}_\mathbf{s}[0] + 1) \div 32 })$ \\
\midrule
0x56 & {\small SLOAD} & 1 & 1 & Load word from storage. \\
&&&& $\mathbb{M}'_\mathbf{s}[0] \equiv \mathbb{S}[I_a]_\mathbf{s}[\mathbb{M}_\mathbf{s}[0]]$ \\
\midrule
0x57 & {\small SSTORE} & 2 & 0 & Save word to storage. \\
&&&& $\mathbb{S}'[I_a]_\mathbf{s}[ \mathbb{M}_\mathbf{s}[0] ] \equiv \mathbb{M}_\mathbf{s}[1] $ \\
&&&& $C_{\text{\tiny SSTORE}}(\mathbb{S}, \mathbb{M}) \equiv \begin{cases}
2G_{sstore} & \text{if} \quad \mathbb{M}_\mathbf{s}[1] \neq 0 \; \wedge \; \mathbb{S}[I_a]_\mathbf{s}[\mathbb{M}_\mathbf{s}[0]] = 0 \\
0 & \text{if} \quad \mathbb{M}_\mathbf{s}[1] = 0 \; \wedge \; \mathbb{S}[I_a]_\mathbf{s}[\mathbb{M}_\mathbf{s}[0]] \neq 0 \\
G_{sstore} & \text{otherwise}
\end{cases}$ \\
\midrule
0x58 & {\small JUMP} & 1 & 0 & Alter the program counter. \\
&&&& $J_{\text{\tiny JUMP}}(\mathbb{M}) \equiv \mathbb{M}_\mathbf{s}[0] $ \\
&&&& This has the effect of writing said value to $\mathbb{M}_{pc}$. See section \ref{ch:model}. \\
\midrule
0x59 & {\small JUMPI} & 2 & 0 & Conditionally alter the program counter. \\
&&&& $J_{\text{\tiny JUMPI}}(\mathbb{M}) \equiv \begin{cases} \mathbb{M}_\mathbf{s}[0] & \text{if} \quad \mathbb{M}_\mathbf{s}[1] = 0 \\ \mathbb{M}_{pc} + 1 & \text{otherwise} \end{cases} $ \\
&&&& This has the effect of writing said value to $\mathbb{M}_{pc}$. See section \ref{ch:model}. \\
\midrule
0x5a & {\small PC} & 0 & 1 & Get the program counter. \\
&&&& $\mathbb{M}'_\mathbf{s}[0] \equiv \mathbb{M}_{pc}$ \\
\midrule
0x5b & {\small MSIZE} & 0 & 1 & Get the size of active memory. \\
&&&& $\mathbb{M}'_\mathbf{s}[0] \equiv \mathbb{M}_{i}$ \\
\midrule
0x5c & {\small GAS} & 0 & 1 & Get the amount of available gas. \\
&&&& $\mathbb{M}'_\mathbf{s}[0] \equiv \mathbb{M}_{g}$ \\
\bottomrule
\end{tabular*}

@@

The header of the genesis block is 9 items, and is specified thus:

$\langle 0_{256}, \mathtt{\tiny SHA3}(\mathtt{\tiny RLP}([])), 0_{160}, 0_{256}, \mathtt{\tiny SHA3}(\mathtt{\tiny RLP}([])), 2^{22}, 0, (), 42 \rangle$

@@

$()$ refers to the extradata, an empty byte array.

$\mathtt{\tiny SHA3}(\mathtt{\tiny RLP}([]))$ values refer to the hashes of the transaction and uncle lists in RLP, both empty.

>>>>

Dealings in this proposed system would have several attributes not often found in the real word. The incorruptibility of judgement, often difficult to find, comes naturally from a disinterested algorithmic interpreter. Transparency, or being able to see exactly how a state or judgement came about through the transaction log and rules or instructional codes, never happens perfectly in human-based systems since natural language is necessarily vague, information is often lacking, and plain old prejudices are difficult to shake.

Overall, I wish to provide a system such that users can be guaranteed that no matter with which other individuals, systems or organisations they interact, they can do so with absolute confidence in the possible outcomes and how those outcomes might come about.

\subsection{Previous Work} \label{ch:previous}

\cite{buterin2013ethereum} first proposed the kernel of this work in late November, 2013. Though now evolved in many ways, the key functionality of a block-chain with a Turing-complete language and an effectively unlimited inter-transaction storage capability remains unchanged.

Hashcash, introduced by \cite{laurie2004proof}, provided the first work into the usage of a cryptographic proof of computational expenditure as a means of transmitting a value signal over the Internet. Though not widely adopted, the work was later utilised and expanded upon by \cite{nakamoto2008bitcoin} in order to devise a cryptographically secure mechanism for coming to a decentralised social consensus over the order and contents of a series of cryptographically signed financial transactions. The fruits of this project, Bitcoin, provided a first glimpse into a decentralised transaction ledger.

Other projects built on Bitcoin's success; the alt-coins introduced numerous other currencies through alteration to the protocol. The best known are Litecoin and Primecoin, discussed by \cite{sprankel2013technical}. Other projects sought to take the core value content mechanism of the protocol and repurpose it; \cite{aron2012bitcoin} discusses, for example, the Namecoin project which aims to provide a decentralised name-resolution system.

Other projects still aim to build upon the Bitcoin network itself, leveraging the large amount of value placed in the system and the vast amount of computation that goes into the consensus mechanism. The Mastercoin project, first proposed by \cite{mastercoin2013willett}, aims to build a richer protocol involving many additional high-level features on top of the Bitcoin protocol though utilisation of a number of auxilliary parts to the core protocol. The Coloured Coins project, proposed by \cite{colouredcoins2012rosenfeld}, takes a similar but more simplified strategy, embellishing the rules of a transactions in order to break the fungibility of Bitcoin's base currency and allow the creation and tracking of tokens through a special ``chroma-wallet'' protocol aware piece of software.

Additional work has been done in the area with discarding the decentralisation foundation; Ripple, discussed by \cite{boutellier2014pirates}, has sought to create a ``federated'' system for currency exchange, effectively creating a new financial clearing system. It has demonstrate that high efficiency gains can be made if the decentralisation premise is discarded.

Early work on smart contracts has been done by \cite{szabo1997formalizing} and \cite{miller1997future}. Around the 1990s it became clear that algorithm enforcement of agreements could become a significant force in human cooperation. Though no specific system was proposed to implement such a system, it was proposed that the future of law would be heavily affected by such systems. In this light, Ethereum may be seen as a general implementation of such a \textit{crypto-law} system.

%E language?

@@

Ethereum, taken as a whole, can be viewed as a transaction-based state machine: we begin with a genesis state and incrementally execute transactions to morph it into some final state. It is this final state which we accept as the canonical ``version'' of the world of Ethereum. The state can include such information as account balances, reputations, trust arrangements, data pertaining to information of the physical world; in short, anything that can currently be represented by a computer is admissable. Transactions thus represent a valid arc between two states; the `valid' part is important---there exist far more invalid state changes than valid state changes. Invalid state changes might, \eg be things such as reducing an account balance without an equal and opposite increase elsewhere. A valid state transition is one which comes about through a transaction. Formally:

@@

Transactions are collated into blocks; blocks are chained together using a cryptographic hash as a means of reference. Blocks function as a journal, recording a series of transactions together with the previous block and an identifier for the final state (though do not store the final state itself---that would be far too big). They also punctuate the transaction series with incentives for nodes to \textit{mine}. This incentivisation takes places as a state-transition function, adding value to a nominated account.

@@

In order to incentivise computation within the network, there needs to be an agreed method for transmitting value. To address this issue, Ethereum has an intrinsic currency, Ether, known also as {\small ETH} and sometimes refered to by the Old English \DH{}. The smallest subdenomination of Ether, and thus the one in which all integer values of the currency are counted, is the Wei. One Ether is defined as being $10^{18}$ Wei. There exist other subdenominations of Ether:
\par`
\begin{center}
\begin{tabular}{rl}
\toprule
Multiplier & Name \\
\midrule
$10^0$ & Wei \\
$10^{12}$ & Szabo \\
$10^{15}$ & Finney \\
$10^{18}$ & Ether \\
\bottomrule
\end{tabular}
\end{center}
\par

@@

We define the checkpoint state $\mathbb{S}_0$:
\begin{eqnarray}
\mathbb{S}_0 & \equiv & \mathbb{S} \quad \text{except:} \\
\mathbb{S}_0[S(T)]_b & \equiv & \mathbb{S}[S(T)]_b - v_0 \\
\mathbb{S}_0[S(T)]_n & \equiv & \mathbb{S}[S(T)]_n + 1
\end{eqnarray}

@@

We define $\mathbb{S}_1$, the checkpoint state as the original state but with the value transferred to the recipient:
\begin{eqnarray}
\mathbb{S}_1 & \equiv & \mathbb{S} \quad \text{except:} \\
\mathbb{S}_1[r]_b & \equiv & \mathbb{S}_1[r]_b + v
\end{eqnarray}

If the recipient account contains no code to be executed (i.e. it's not a contract), then we define the final state as being equivalent to the checkpoint state. In the case that the recipient contains code to be executed (i.e. that the account is a contract), then the contract's body code (identified as the fragment whose SHA3 hash is $\mathbb{S}[r]_c$) is executed according to the execution model (see section \ref{ch:model}). Just as with contract creation, if the execution halts due to an exhausted gas supply, then no gas is refunded to the caller and the state is reverted to the point immediately prior to code execution.
\begin{eqnarray}
\mathbb{S}_P & \equiv & \begin{cases}
\mathbb{S}_1 & \text{if} \quad \mathbb{S}_1[r]_c = \varnothing \\
\mathbb{S}_1 & \text{if} \quad \mathbb{S}^{**} = \varnothing \\
\mathbb{S}^{**} & \text{otherwise}
\end{cases} \\
(\mathbb{S}^{**}, g', \mathbf{o}) & \equiv & \Xi(\mathbb{S}_1, g, I) \\
I_a & \equiv & a \\
I_o & \equiv & o \\
I_p & \equiv & p \\
I_\mathbf{d} & \equiv & \mathbf{d} \\
I_s & \equiv & s \\
I_v & \equiv & v \\
\text{Let} \quad \mathtt{\tiny SHA3}(I_\mathbf{b}) & = & \mathbb{S}[r]_c
\end{eqnarray}

@@

The state database won't be forced to maintain all past state trie structures in to the future. It should maintain an age for each node and eventually discard nodes that are neither recent enough nor checkpoints; checkpoints, or a set of nodes in the database that allow a particular block's state trie to be traversed, could be used to place a maximum limit on the amount of computation needed in order to retrieve any state throughout the blockchain.

Blockchain consolidation could be used in order to reduce the amount of blocks a client would need to download to act as a full, mining, node. A compressed archive of the trie structure at given points in time (perhaps one in every 10000th block) could be maintained by the peer network, effectively recasting the genesis block. This would reduce the amount to be downloaded to a single archive plus a hard maximum limit  of blocks.

Finally, blockchain compression could perhaps be conducted: nodes in state trie that haven't sent/received a transaction in over some constant amount of blocks could be thrown out, reducing both Ether-leakage and the growth of the state database.

\subsection{Scalability}

Scalability remains an eternal concern. With a generalised state transition function, it becomes difficult to partition and parallelise transactions to apply the divide-and-conquer strategy. Unaddressed, the dynamic value-range of the system remains essentially fixed and as the average transaction value increases, the less valuable of them become ignored, being economically pointless to include in the main ledger. However, several strategies exist that may potentially be exploited to provide a considerably more scalable protocol.

Some form of hierarchical structure, either through consolidating smaller lighter-weight chains into the main block or through building the main block through the incremental combination and adhesion (through proof-of-work) of smaller transaction sets may allow parallelisation of transaction combination and block-building. Parallelism could also come from a prioritised set of parallel blockchains, consolidated each block and with duplicate or invalid transactions thrown out accordingly.

Verifyable computation, if made generally available and efficient enough, may provide a route to allow the proof-of-work to be the verification of final state.

@@

I have introduced, discussed and formally defined the protocol of Ethereum. Through this protocol the reader may implement a node on the Ethereum network and join others in a decentralised secure social operating system. Contracts may be authored in order to algorithmically specify and autonomously enforce rules of interaction.

@@

Useful corrections and suggestions were provided by a number of others from the Ethereum community including Aeron Buchanan, Nick Savers, Viktor Tr\'{o}n, Marko Simovic and, of course, Vitalik Buterin.

@@

\section{Recursive Length Prefix}\label{app:rlp}
This is a serialisation method for encoding arbitrarily structured binary data (byte arrays).

@@

If RLP is used to encode a positive integer, it must be specified as the shortest byte array such that the big-endian interpretation of it is equal. Thus the RLP of some positive integer $i$ is defined as:
\begin{equation}
\mathtt{\tiny RLP}(i : i \in \mathbb{P}) \equiv \mathtt{\tiny RLP}(\mathtt{\tiny BE}(i))
\end{equation}

There is no specific canonical encoding format for signed or floating-point values.

@@

\begin{eqnarray}
\mathtt{\tiny HP}(\mathbf{x}, t): \mathbf{x} \in \mathbb{E} & \equiv & \begin{cases}
(16f(t), 16\mathbf{x}[0] + \mathbf{x}[1], 16\mathbf{x}[2] + \mathbf{x}[3], ...) &
\text{if} \quad \lVert \mathbf{x} \rVert \; \text{is even} \\
(16(f(t) + 1) + \mathbf{x}[0], 16\mathbf{x}[1] + \mathbf{x}[2], 16\mathbf{x}[3] + \mathbf{x}[4], ...) &
\text{otherwise}
\end{cases} \\
f(t) & \equiv & \begin{cases} 2 & \text{if} \quad t \\ 0 & \text{otherwise} \end{cases}
\end{eqnarray}

Thus the high nibble of the first byte contains two flags; the lowest bit encoding the oddness of the length and the second-lowest encoding the flag $t$. The low nibble of the first byte is zero in the case of an even number of nibbles and the first nibble in the case of an odd number. All remaining nibbles (now an even number) fit properly into the remaining bytes.

\section{Modified Merkle Patricia Tree}\label{app:trie}
The modified Merkle Patricia tree (trie) provides a persistent data structure to map between arbitrary-length binary data (byte arrays). It is defined in terms of a mutable data structure to map between 256-bit binary fragments and arbitrary-length binary data, typically implemented as a database. The core of the trie, and its sole rquirement in terms of the protocol specification is to provide a single 32-byte value that identifies a given set of key-value pairs. It is left as an implementation consideration to store and maintain the structure of the trie in a manner the allows effective and efficient realisation of the protocol.

Formally, we assume the input value $\mathfrak{I}$, a set containing pairs of byte sequences:
\begin{equation}
\mathfrak{I} = \{ (\mathbf{k}_0 \in \mathbb{B}, \mathbf{v}_0 \in \mathbb{B}), (\mathbf{k}_1 \in \mathbb{B}, \mathbf{v}_1 \in \mathbb{B}), ... \}
\end{equation}

When considering such a sequence, we use the common numeric subscript notation to refer to a tuple's key or value, thus:
\begin{equation}
\forall_{I: \in \mathfrak{I}} I \equiv (I_0, I_1)
\end{equation}

Any series of bytes may also trivially be viewed as a series of nibbles, given an endian-specific notation; here we assume big-endian. Thus:
\begin{eqnarray}
y(\mathfrak{I}) & = & \{ (\mathbf{k}_0' \in \mathbb{Y}, \mathbf{v}_0 \in \mathbb{B}), (\mathbf{k}_1' \in \mathbb{Y}, \mathbf{v}_1 \in \mathbb{B}), ... \} \\
\forall_n \quad \forall_{i: i < 2\lVert\mathbf{k}_n\rVert} \quad \mathbf{k}_n'[i] & \equiv &
\begin{cases}
\lfloor \mathbf{k}_n[\lfloor i \div 2 \rfloor] \div 16 \rfloor & \text{if} \; i \; \text{is even} \\
\mathbf{k}_n[\lfloor i \div 2 \rfloor] \bmod 16 & \text{otherwise}
\end{cases}
\end{eqnarray}

We define the function $\texttt{\small TRIE}$, which evaluates to the root of the trie that represents this set when encoded in this structure. The empty trie is defined as being a 256-bit hash of zeroes, $0_{256}$:`
\begin{equation}
\texttt{\small TRIE}(\mathfrak{I}) \equiv \begin{cases}
0_{256} & \text{if} \quad \mathfrak{I} = \varnothing \\
n(y(\mathfrak{I}), 0) & \text{otherwise}
\end{cases}
\end{equation}

We assume a function $n$, the trie's node cap function. When composing a node, we use RLP to encode the structure. As a means of reducing storage complexity, for nodes whose composed RLP is fewer than 32 bytes, we store the RLP directly; for those larger we assert prescience of the byte array whose SHA3 hash evaluates to our reference. Thus we define in terms of $c$, the node composition function:
\begin{equation}
n(\mathfrak{I}, i) \equiv \begin{cases}
() & \text{if} \quad \mathfrak{I} = \varnothing \\
c(\mathfrak{I}, i) & \text{if} \quad \lVert c(\mathfrak{I}, i)\rVert < 32 \\
\texttt{\small SHA3}(c(\mathfrak{I}, i)) & \text{otherwise}
\end{cases}
\end{equation}

In a manner similar to a radix tree, when the trie is traversed from root to leaf, one may build a single key-value pair. The key is accumulated through the traversal, acquiring a single nibble from each branch node (just as with a radix tree). Unlike a radix tree, in the case of multiple keys shared the same prefix or in the case of a single key having a unique suffix, two optimising nodes are provided. Thus while traversing, one may potentially acquire multiple nibbles from each of the other two node types, extension and leaf. There are three kinds of nodes in the trie:
\begin{description}
\item[Leaf] A two-item structure whose first item corresponds to the nibbles in the key not already accounted for by the accumulation of keys and branches traversed from the root. The hex-prefix encoding method is used and the second parameter to the function is required to be $true$.
\item[Extension] A two-item structure whose first item corresponds to a series of nibbles of size greater than one that are shared by at least two distinct keys past the accumulation of nibbles keys and branches as traversed from the root. The hex-prefix encoding method is used and the second parameter to the function is required to be $false$.
\item[Branch] A 17-item structure whose first sixteen items correspond to each of the sixteen possible nibble values for the keys at this point in their traversal. The 17th item is used in the case of this being a terminator node and thus a key being ended at this point in its traversal.
\end{description}

A branch is then only used when necessary; no branch nodes may exist that contain only a single non-zero entry. We may formally define this structure with the structural composition function $c$:
\begin{equation}
c(\mathfrak{I}, i) \equiv \begin{cases}
\texttt{\small RLP}\Big( \big(\texttt{\small HP}(I_0[i .. (\lVert I_0\rVert - 1)], I_1 \big) \Big) & \text{if} \quad \lVert \mathfrak{I} \rVert = 1 \quad \text{where} \; \exists I: I \in \mathfrak{I} \\
\texttt{\small RLP}\Big(\big(\texttt{\small HP}(I_0[i .. (j - 1)]), n(\mathfrak{I}, j) \big) \Big) & \text{if} \quad i \ne j \quad \text{where} \; j = \arg \max_x : \exists \mathbf{l}: \lVert \mathbf{l} \rVert = x : \forall_{I \in \mathfrak{I}}: I_0[0 .. (x - 1)] = \mathbf{l} \\
\texttt{\small RLP}\Big( (u(0), u(1), ..., u(15), v) \Big) & \text{otherwise} \quad \text{where} \begin{array}[t]{rcl}
u(n) & \equiv & n(\{ I : I \in \mathfrak{I} \wedge I_0[i] = n \}, i + 1) \\
v & = & \begin{cases}
I_1 & \text{if} \quad \exists I: I \in \mathfrak{I} \wedge \lVert I_0 \rVert = i \\
() & \text{otherwise}
\end{cases}
\end{array}
\end{cases}
\end{equation}

\subsection{Trie Database}
Thus no explicit assumptions are made concerning what data is stored and what is not, since that is an implementation-specific consideration; we simply define the identity function mapping the key-value set $\mathfrak{I}$ to a 32-byte hash and assert that only a single such hash exists for any $\mathfrak{I}$, which though not strictly true is accurate within acceptable precision. In reality, a sensible implemention will not fully recompute the trie root hash for each set.

A reasonable implementation will maintain a database of nodes determined from the computation of various tries or, more formally, it will memoise the function $c$. This strategy uses the nature of the trie to both easily recall the contents of any previous key-value set and to store multiple such sets in a very efficient manner. Due to the dependency relationship, Merkle-proofs may be constructed with an $O(\log N)$ space requirement that can demonstrate a particular leaf must exist within a trie of a given root hash.

@@

@@

The method of signing transactions is similar to the `Electrum style signatures'; it utilises the SECP-256k1 curve as described by \cite{gura2004comparing}. This has yet to be formally specified.

@@

\begin{tabular*}{\columnwidth}[h]{rlrrl}
\toprule
\multicolumn{5}{c}{\textbf{50s: Stack, Memory, Storage and Flow Operations}} \vspace{5pt} \\
\textbf{Value} & \textbf{Mnemonic} & $\delta$ & $\alpha$ & \textbf{Description} \vspace{5pt} \\
0x50 & {\small POP} & 1 & 0 & Remove item from stack. \\
\midrule
0x51 & {\small DUP} & 1 & 2 & Duplicate stack item. \\
&&&& $\mathbb{M}'_\mathbf{s}[0] \equiv \mathbb{M}_\mathbf{s}[0]$ \\
\midrule
0x52 & {\small SWAP} & 2 & 2 & Exchange stack items. \\
&&&& $\mathbb{M}'_\mathbf{s}[0] \equiv \mathbb{M}_\mathbf{s}[1]$ \\
&&&& $\mathbb{M}'_\mathbf{s}[1] \equiv \mathbb{M}_\mathbf{s}[0]$ \\
\midrule
0x53 & {\small MLOAD} & 1 & 1 & Load word from memory. \\
&&&& $\mathbb{M}'_\mathbf{s}[0] \equiv \mathbb{M}_\mathbf{m}[\mathbb{M}_\mathbf{s}[0] \dots (\mathbb{M}_\mathbf{s}[0] + 31) ]$ \\
&&&& $\mathbb{M}'_i \equiv \max(\mathbb{M}_i, \ceil{ (\mathbb{M}_\mathbf{s}[0] + 32) \div 32 })$ \\
\midrule
0x54 & {\small MSTORE} & 2 & 0 & Save word to memory. \\
&&&& $\mathbb{M}'_\mathbf{m}[ \mathbb{M}_\mathbf{s}[0] \dots (\mathbb{M}_\mathbf{s}[0] + 31) ] \equiv \mathbb{M}_\mathbf{s}[1]$ \\
&&&& $\mathbb{M}'_i \equiv \max(\mathbb{M}_i, \ceil{ (\mathbb{M}_\mathbf{s}[0] + 32) \div 32 })$ \\
\midrule
0x55 & {\small MSTORE8} & 2 & 0 & Save byte to memory. \\
&&&& $\mathbb{M}'_\mathbf{m}[ \mathbb{M}_\mathbf{s}[0] ] \equiv (\mathbb{M}_\mathbf{s}[1] \bmod 256) $ \\
&&&& $\mathbb{M}'_i \equiv \max(\mathbb{M}_i, \ceil{ (\mathbb{M}_\mathbf{s}[0] + 1) \div 32 })$ \\
\midrule
0x56 & {\small SLOAD} & 1 & 1 & Load word from storage. \\
&&&& $\mathbb{M}'_\mathbf{s}[0] \equiv \mathbb{S}[I_a]_\mathbf{s}[\mathbb{M}_\mathbf{s}[0]]$ \\
\midrule
0x57 & {\small SSTORE} & 2 & 0 & Save word to storage. \\
&&&& $\mathbb{S}'[I_a]_\mathbf{s}[ \mathbb{M}_\mathbf{s}[0] ] \equiv \mathbb{M}_\mathbf{s}[1] $ \\
&&&& $C_{\text{\tiny SSTORE}}(\mathbb{S}, \mathbb{M}) \equiv \begin{cases}
2G_{sstore} & \text{if} \quad \mathbb{M}_\mathbf{s}[1] \neq 0 \; \wedge \; \mathbb{S}[I_a]_\mathbf{s}[\mathbb{M}_\mathbf{s}[0]] = 0 \\
0 & \text{if} \quad \mathbb{M}_\mathbf{s}[1] = 0 \; \wedge \; \mathbb{S}[I_a]_\mathbf{s}[\mathbb{M}_\mathbf{s}[0]] \neq 0 \\
G_{sstore} & \text{otherwise}
\end{cases}$ \\
\midrule
0x58 & {\small JUMP} & 1 & 0 & Alter the program counter. \\
&&&& $J_{\text{\tiny JUMP}}(\mathbb{M}) \equiv \mathbb{M}_\mathbf{s}[0] $ \\
&&&& This has the effect of writing said value to $\mathbb{M}_{pc}$. See section \ref{ch:model}. \\
\midrule
0x59 & {\small JUMPI} & 2 & 0 & Conditionally alter the program counter. \\
&&&& $J_{\text{\tiny JUMPI}}(\mathbb{M}) \equiv \begin{cases} \mathbb{M}_\mathbf{s}[0] & \text{if} \quad \mathbb{M}_\mathbf{s}[1] = 0 \\ \mathbb{M}_{pc} + 1 & \text{otherwise} \end{cases} $ \\
&&&& This has the effect of writing said value to $\mathbb{M}_{pc}$. See section \ref{ch:model}. \\
\midrule
0x5a & {\small PC} & 0 & 1 & Get the program counter. \\
&&&& $\mathbb{M}'_\mathbf{s}[0] \equiv \mathbb{M}_{pc}$ \\
\midrule
0x5b & {\small MSIZE} & 0 & 1 & Get the size of active memory. \\
&&&& $\mathbb{M}'_\mathbf{s}[0] \equiv \mathbb{M}_{i}$ \\
\midrule
0x5c & {\small GAS} & 0 & 1 & Get the amount of available gas. \\
&&&& $\mathbb{M}'_\mathbf{s}[0] \equiv \mathbb{M}_{g}$ \\
\bottomrule
\end{tabular*}

@@

The genesis block is 9 items, and is specified thus:

$\big( \big( 0_{256}, \mathtt{\tiny SHA3}\big(\mathtt{\tiny RLP}\big( () \big)\big), 0_{160}, 0_{256}, \mathtt{\tiny SHA3}\big(\mathtt{\tiny RLP}\big( () \big)\big), 2^{22}, 0, (), 42 \big), (), () \big)$

@@

$()$ refers to the extradata and the sequences of both uncles and transactions, all empty.

$\mathtt{\tiny SHA3}\big(\mathtt{\tiny RLP}\big( () \big)\big)$ values refer to the hashes of the transaction and uncle lists in RLP, both empty.
